---
title: An Introduction to `Rclusterpp`
author: Michael Linderman
bibliography: bibtex/ref.bib
date: '`r Sys.Date()`'
output:
  html_document:
  fig_caption: yes
  theme: cerulean
  toc: yes
  toc_depth: 2
vignette: >
  %\VignetteIndexEntry{An Introduction to `Rclusterpp`}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

## Introduction

Hierarchical clustering is a fundamental data analysis tool. However,
the $O(n^2)$ memory footprint of commonly available implementations,
such as `stats::hclust`, which maintain the dissimilarity matrix in
memory (colloquially stored-distance) limit these implementations to
tens of thousands of observations or less. In the motivating domain for
this work, flow cytometry, datasets are hundreds of thousands or even
millions of observations in size (but with comparatively low
dimensionality, e.g., less than 30). In this and other similar contexts
building out the complete distance matrix is not possible and
alternative implementations with $O(n)$ memory footprint are needed.

The memory requirements of hierarchical clustering have motivated the
development of alternative clustering algorithms that do not require the
full dissimilarity matrix. Such algorithms are not the focus of
Rclusterpp. Instead we focus on the common situation wherein a complex
data analysis pipeline, which includes hierarchical clustering, is first
designed and validated on smaller datasets, and only later scaled to
larger in inputs. In these cases, we wish to maintain the same
functionality, an if possible the same results, but scale efficiently.
Thus the goal for Rclusterpp is to provide efficient "stored data"
implementations for common hierarchical clustering routines, e.g.,
single, complex, average and Ward's linkage, that scale to hundreds of
thousands of observations while aiming to deliver results identical to
the "stock" `stats::hclust` implementation (if the cluster hierarchy is
unambiguous, then the results should be identical).

As an example, the following two statements produce identical results:

```{r}
library(Rclusterpp)
h <- hclust(dist(USArrests, method="euclidean"), method="average")
r <- Rclusterpp.hclust(USArrests, method="average", distance="euclidean")
# Check equality of the dedrogram tree and agglomeration heights
identical(h$merge, r$merge) && all.equal(h$height, r$height)
```

however, in the latter, the memory footprint is on the order of $O(n)$
as opposed to $O(n^2)$, for $n$ observations (ignoring the footprint
of the data itself). When required, such as in the example above,
Rclusterpp purposely trades time for space to maintain a $O(n)$ memory
footprint. Section @sec:data includes a summary of the complexity of
each linkage method as implemented.

The computational demanding components of Rclusterpp are implemented
in [C++]{.sans-serif} using OpenMP ^[OpenMP is only enabled on Linux
and OSX due to issues with the pthreads compatibility DLL on Windows.]
to take advantage of multi-core processors and multi-processor shared
memory computers. Thus even when incurring additional computation
costs to reduce the memory footprint Rclusterpp is faster than
`stats::hclust`, and in cases, such as Ward's linkage, where no such
trade-off exists, Rclusterpp can be faster than even the "fast"
stored-distance clustering packages like fastcluster.  Sample
benchmark results are shown in Table 1.

```{r, echo = FALSE}
d  <- scan(textConnection("
  Rclusterpp            0.0006    100
  fastcluster           0.0008    100
  hclust                0.0028    100
  Rclusterpp            0.0036    500
  fastcluster           0.0208    500
  hclust                0.2606    500
  Rclusterpp            0.0092    1000
  fastcluster           0.1252    1000
  hclust                1.8012    1000
  Rclusterpp            0.1814    5000
  fastcluster           2.6246    5000
  hclust               199.1894   5000"),
  character(0))
d  <- as.data.frame(matrix(d, ncol = 3, byrow = TRUE), stringsAsFactors = FALSE)
d$V2  <- as.numeric(d$V2); d$V3  <- as.integer(d$V3)
names(d)  <- c("Implementation", "Exec. Time (s)", "$n$")
knitr::kable(d, caption = "Table 1: Execution time averaged across 5 runs for various clustering implementations (including distance computation) for Ward's minimum variance method for $n\times 10$ input data measured on a quad-core 3.05 GHz Intel i7 950 server."
)
```

In some applications, such as the WGCNA [@Zhang2005] algorithm that also
motivated this work, the dissimilarity matrix is already computed in a
previous stage of the workflow and thus there is no advantage to be
gained with stored-data approaches. However, memory footprint is still a
concern. Those individuals who have attempted to cluster more than 46340
observations have discovered that [R]{.sans-serif} limits matrices to
$2^31$ elements or less. In these cases, it is desirable to perform all
of the memory intensive components of the workflow on the
"[C++]{.sans-serif} side", where there is no such limit and where the
implementor has more control over the creation of temporaries.
Rclusterpp exposes its various clustering implementations as a templated
library that can be linked against by other [C++]{.sans-serif}-backed R
packages (modeled on the techniques used in the Rcpp package).

## Stored-data Hierarchical Clustering {#sec:data}

`Rclusterpp` currently implements a subset of the clustering methods and
distance metrics provided by `stats::hclust`. Specifically, Rclusterpp
currently supports the following linkage methods:

```{r}
Rclusterpp.linkageKinds()
```

and the following distance metrics:

```{r}
Rclusterpp.distanceKinds()
```

The linkage methods are currently limited to reducible geometric methods
that can implemented exactly using the *recursive nearest neighbor
(RNN)* algorithm [@Murtagh1983].

Table 2 shows the estimated worst-case time and
space complexities [@Murtagh1984] for the algorithms used in `Rclusterpp`.
Ward's and single-link are implemented with optimal time and space using
the RNN and SLINK [@Sibson1973] algorithms respectively; while average
and complete-link trade increased time bounds, in exchange for reducing
the memory footprint to $O(n)$ from $O(n^2)$.

```{r, echo = FALSE}
d  <- scan(textConnection("
  Average        RNN        $O(n^3*m)$           $O(n)$
  Complete       RNN        $O(n^3*m)$           $O(n)$
  Ward           RNN        $O(n^2*m)$          $O(n*m)$
  Single        SLINK       $O(n^2*m)$           $O(n)$
"), character(0))
d  <- as.data.frame(matrix(d, ncol = 4, byrow = TRUE), stringsAsFactors = FALSE)
names(d) <- c("Method", "Algorithm", "Time Complexity", "Space Complexity")
knitr::kable(d, caption = "Table 2: Worst-case time and space complexities for the Rclusterpp
  stored-data implementation (not including the original $O(n*m)$ data
  footprint)")
```

As shown previously, the `Rclusterpp.hclust` function has a very similar
interface to `stats::hclust`, but will also accept a numeric matrix
(instead of a `dist` object) and a distance metric. The return value is
the same `hclust` object as produced by `stats::hclust`.

Since the underlying components of the clustering implementation,
including the RNN implementation, linkage methods and distance
functions, are all exposed as a templated C++ library, users can readily
create derivative packages that implement custom clustering
methodologies without starting from scratch.
Section @sec:distance has more information on how to work with the
C++ library (in the context of stored-distance implementations, but the
information is just as a applicable to stored-data). In addition, the
interested user is pointed to the source for `hclust_from_data`, the
[C++]{.sans-serif} function called by `Rclusterpp.hclust`, which is
itself a consumer of the templated library.

## Stored-distance Hierarchical Clustering (in C++) {#sec:distance}

`Rclusterpp.hclust` can be used as a limited-functionality replacement
for `stats::hclust`, i.e., it will accept a `dist` object as input.
However, as shown Table 3, in this usage Rcpp is often slower than
fastcluster and other packages specifically optimized for this use case.
Instead, Rclusterpp's stored-distance functionality is intended for use
as linkable C++ library.

```{r, echo = FALSE}
d  <- scan(textConnection("
  fastcluster               0.0008    100
  RclusterppDistance        0.0014    100
  Rclusterpp                0.0014    100
  hclust                    0.0028    100
  fastcluster               0.0182    500
  RclusterppDistance        0.0220    500
  Rclusterpp                0.0376    500
  hclust                    0.2354    500
  fastcluster               0.1306    1000
  RclusterppDistance        0.1364    1000
  Rclusterpp                0.1508    1000
  hclust                    1.7396    1000
  fastcluster               2.4642    5000
  RclusterppDistance        2.8008    5000
  Rclusterpp                5.5344    5000
  hclust                   204.2906   5000
"), character(0))
d  <- as.data.frame(matrix(d, ncol = 3, byrow = TRUE), stringsAsFactors = FALSE)
d$V2  <- as.numeric(d$V2); d$V3  <- as.integer(d$V3)
names(d)  <- c("Implementation", "Exec. Time (s)", "$n$")
knitr::kable(d, caption = "Table 3: Execution time averaged across 5 runs for various clustering implementations (including distance computation) for average-link/euclidean distance clustering on $n\times 10$ input data measured on a quad-core 3.05 GHz Intel i7 950 server. `RclusterppDistance` is the stored-distance implementation and `Rclusterpp` is the stored data implementation.")
```

Rclusterpp is modeled on the Rcpp\* family of packages. Rclusterpp
provides its own skeleton function, `Rclusterpp.package.skeleton`, which
can be used to generate new packages that are setup to link against the
Rclusterpp library. Alternately one can use the inline package to
compile C++ code from within R. The Rclusterpp package includes an
example "inline\" function, shown below, which we will use as our
working example in this document.

```{bash, code = readLines(system.file("examples","clustering.R",package="Rclusterpp")), eval = FALSE}
```

Rclusterpp makes extensive use of Rcpp to build the interface between
[R]{.sans-serif} and [C++]{.sans-serif}, and the Eigen library (via
RcppEigen) for matrix and vector operations. A working knowledge of both
libraries will be needed to effectively use Rclusterpp as this lower
level.

Rclusterpp provides several convenience `typedef`s for working at the
interface of Eigen and [R]{.sans-serif}, in this case, we use
`MapNumericMatrix` to wrap, or "map", an Eigen `Matrix` around the
[R]{.sans-serif} data pointer (and thus no copy is involved) for use
with Eigen operators. We further create a `NumericMatrix` to store the
distance matrix we will compute, and extract a reference to the strictly
lower portion of that matrix for use in the clustering routine.

At present, Rclusterpp assumes the dissimilarities are in the strictly
lower portion of the matrix, and will not work with other inputs.

Agglomerations are tracked in `ClusterVector` object. The user needs to
select the appropriate cluster type for their problem. A templated type
factory, `ClusterTypes` is provided to assist in this selection
(`NumericCluster` is a convenience `typedef` for this factory for
`NumericMatrix`).
```{Rcpp, eval = FALSE}
typedef ClusterTypes<Rcpp::NumericMatrix::stored_type> NumericCluster;  
NumericCluster::plain;  // Simplest cluster, used for stored-distance
NumericCluster::center; // Maintains cluster "center", used for Ward's linkage
NumericCluster::obs;    // Tracks obs in each cluster, used for Average, Complete...
```

Clustering is performed by specifying the clustering method, i.e., RNN,
the linkage method and the initialized cluster vector. In this case we
are performing stored-distance average link clustering using the
distance matrix computed earlier. Note that the stored-distance linkage
methods are implemented with Lance-Williams update algorithm and are
destructive to the strictly lower portion of the dissimilarity matrix.

At the completion of the clustering, the cluster vector will contain all
of the agglomerations along with the agglomeration heights. Rclusterpp
extends Rcpp with a `wrap` implementation that will translate that
vector into a [R]{.sans-serif} list with the `merge`, `height` and
`order` entries needed for the `hclust` object.

## References
